---
title: "Lesson 4: Understanding Missing Values"
subtitle: "What does it mean when data are missing?"
---

```{r}
#| include: false
source("../data/lock5-usstates.R")
```

## Learning Objectives

By the end of this lesson, you should be able to:

- Explain why `NA` means **unknown**, not zero or blank.
- Predict how missing values affect arithmetic and summary statistics.
- Identify missing values using `is.na()`.
- Count the number and proportion of missing values.
- Remove missing values when appropriate using logical indexing.

[Download the companion R guide for this lesson](found04-missing-values-companion.R)

---

## Introduction

Real datasets are rarely complete. Sometimes a value is missing because it was never collected, a participant skipped a question, or the information could not be determined.

In R, missing values are represented by **`NA`**, which stands for **Not Available**.

The important idea is that **`NA` means "unknown."** It does **not** mean zero or an empty value.

---

## Arithmetic with Missing Values

Suppose we have a vector of quiz scores for a small class, but one student was absent that day. 

The student's grade is not 0; it is missing. They simply haven't made up the quiz yet, so we don't know what their grade will be yet.

```{r}
quiz_scores <- c(8, 10, NA, 7, 9)

quiz_scores

NA + 5
mean(quiz_scores)
```

::: {.callout-note title="Key Idea"}
Missing values propagate uncertainty.

If part of a calculation is unknown, the result is usually unknown.
:::

---


## Finding Missing Values

Let's return to the US States data. Recall that the `fruits` vector measures the percentage of people in each state who eat fruit at least once per day. 

But did every state actually record this data? Or did some states not report their fruit-eating habits? 

```{r}
fruit # hard to read
length(fruit) # not helpful

# check for missing data directly:
is.na(fruit)
```

::: {.callout-note title="Key Idea"}
`is.na()` creates a logical vector.

`TRUE` means the value is missing.

`FALSE` means the value is present.
:::

---

## Counting Missing Values

Once we have a logical vector telling us which values are missing (TRUE) or not missing (FALSE), we can work with it the same way we worked with logical vectors in the last section. 

One of the most common tasks is to count the number of missing (or non-missing) values. This is a great diagnostic to run on any variable. 

```{r}
# How many states are missing the fruit data? 
sum(is.na(fruit))

```


The `physical` variable measures the percent of people in each state who report doing 150+ minutes of aerobic physical activity per week.

Can you adapt the earlier code to count the number of states that did not report their physical activity data? 

```{r}
# How many states are missing the physical activity data? 
#
```

Counting the number of observations that are missing is a good start, but in some datasets, it's more helpful to find the proportion of data that is missing. Is 10% missing? or 40% missing? 
This can make a big difference for whether we are able to use the variable at all. 

Think about how we calculate proportions: PART / WHOLE. In other words:

$$\frac{\text{Number of states with missing data}}{\text{Total number of states}}$$
Can we use R functions we've already learned to calculate each of these pieces? 

Yes!


```{r}

# Number of states with missing physical data:
sum(is.na(physical))

# Total number of states in the dataset: 
length(physical)

# What proportion of states are missing the physical activity data?
sum(is.na(physical)) / length(physical)
mean(is.na(physical))
```

**Shortcut:** remember that R records TRUE as 1 and FALSE as 0. Because of this, we can actually use the mean() function to find the proportion of missing data. Weird, but true! 

$$
\begin{aligned}
\text{mean(is.na(physical))}
    &= \frac{\text{FALSE} + \text{FALSE} + \cdots + \text{TRUE}+ \cdots + \text{FALSE}}{\text{total number of states}} \\[0.8em]
    &= \frac{0 + 0 + \cdots + 1+\cdots+0}{\text{total number of states}} \\[0.8em]
    &= \frac{\text{number of missing values}}{\text{total number of states}} \\[0.8em]
    &= \text{proportion of missing values}
\end{aligned}
$$

**Pro-tip:** Using `mean(is.na(x))` to find the proportion of missing values is a common shortcut among data scientists/statisticians who are too lazy to write out the full calculation. But if you aren't comfortable with why it works, there's no harm in actually writing out `sum(is.na(x)) / length(x)`. Do what makes sense to you. 

---

## Selecting Non-Missing Observations
Sometimes we want to ignore the missing values and work only with the observations that contain data.

To do that, we first need a logical vector that identifies the observed values rather than the missing ones.

The ! operator means NOT. It reverses each logical value:

- TRUE becomes FALSE
- FALSE becomes TRUE

So if `is.na(fruit)` asks "Is this value missing?"

then `!is.na(fruit)` asks "Is this value *not* missing?"

```{r}
is.na(fruit)
!is.na(fruit)
```

Now we have a logical vector with TRUE wherever the data are available.

Just as in the previous lesson, we can use that logical vector inside square brackets ([]) to select the observations we want to keep.

```{r}
fruit_noNA <- fruit[!is.na(fruit)]
fruit_noNA
```

Notice that fruit_noNA contains fewer values than the original fruit vector because the missing observation has been removed.

```{r}
length(fruit)
length(fruit_noNA)
```

In general, be cautious about just removing NA values-- this can end up backfiring, badly. It's much better to investigate cases with missing data and try to discover why the data is missing. Sometimes you can even *impute* (or guess/estimate) missing values.

In practice, we don't often need to manually remove values from a vector, because R has smart options for handling missing data inside calculations (more on that next!). 

If you do decide to manually remove the missing data (only after clearing it with your advisor/boss/professor), then make sure you remove the same cases from any other affected vectors, such as the vector of state names, or the data will no longer match up. 

```{r}
states_noNA <- states[!is.na(fruit)]

head(states_noNA, 10)

length(states_noNA)
length(fruit_noNA)
```



::: {.callout-note title="Key Idea"}
When you remove observations from one vector, you should usually remove the same observations from every related vector so they continue to describe the same observations.
:::

---

## Comparing Summary Statistics

If you try to compute functions or summary statistics on a vector with missing data, R will always report NA as the answer. 

```{r}
mean(fruit)
mean(quiz_scores)
```

That's because, since at least one value is missing, we don't actually know the true mean of the data set.

R forces us to acknowledge this and make a conscious choice about how to handle the missing data. 

If you decide you want to compute the average for the data that is present, then you can. 

```{r}
mean(fruit_noNA)
mean(fruit[!is.na(fruit)])
```

This is a little clunky, and thankfully R has a streamlined option. If you know you have missing data and you want R to remove the missing data and calculate the summary statistic ignoring it, you can: 

```{r}
mean(fruit, na.rm = TRUE)
max(fruit, na.rm = TRUE)
sd(fruit, na.rm = TRUE)
# etc.
```

In this case, `na.rm = TRUE` tells R to set the "remove NA values" option to TRUE. In other words, remove the NA values before calculating the mean. 

Using `na.rm = TRUE` doesn't change the original fruit vector; it only affects the calculation itself.  

---

## Try It

Open the companion R script and investigate the `physical` vector.

1. Which values are missing?
2. How many states are missing physical activity data?
3. What proportion of states are missing physical activity data?
4. Which states are missing physical activity data?
5. What is the mean physical activity value after removing missing values?

---

## Looking Ahead

Missing values are common in real datasets.

Before beginning an analysis, it is good practice to check whether data are missing and think carefully about how those missing values should be handled.

Later in the course, we'll see how these same ideas apply to entire data frames.

---

## Lesson Summary

- `NA` represents an unknown value.
- Missing values usually cause calculations to return `NA`.
- `is.na()` identifies missing values.
- Missing values can be counted with `sum()` or `mean()`.
- Logical indexing with `!is.na()` keeps observed values.
- Related vectors should be filtered together.

---

## R Reference

| Function / Syntax | Purpose |
|-------------------|---------|
| `is.na(x)` | Identify missing values |
| `!is.na(x)` | Identify non-missing values |
| `sum(is.na(x))` | Count missing values |
| `mean(is.na(x))` | Proportion of missing values |
| `mean(x, na.rm = TRUE)` | Ignore missing values when calculating the mean |
