---
title: "Lesson 5: Exploring a Variable"
subtitle: "How can we start exploring a variable?"
---

```{r}
#| include: false
source("../data/lock5-usstates.R")
```

## Learning Objectives

By the end of this lesson, you should be able to:

- Choose an appropriate summary for a quantitative or categorical variable.
- Describe the center and spread of a quantitative variable.
- Identify the categories present in a categorical variable.
- Produce simple frequency tables and bar plots.
- Explain why graphs are an important part of exploratory data analysis.

[Download the companion R guide for this lesson](found05-summarize-variable-companion.R)

---

## Introduction

In the first four lessons, you learned how R stores data, how to create vectors, how to select observations, and how to work with missing values.

Now we're ready to ask an important question:

> **What does this variable look like?**

This process is called **exploratory data analysis (EDA)**.

Rather than jumping directly into formal analyses, we first try to understand our data by calculating simple summaries and creating simple graphs.

---

## Exploring a Quantitative Variable

Suppose we're interested in the percentage of adults who report eating vegetables every day. This is recorded in the `vegetables` vector. 

What can we say about the distribution of vegetable eaters? Let's start with some basic explorations. 

> **How many observations do we have?**

```{r}
length(vegetables)
```

OK good, we have one value for each of the 50 states. But wait...

> **Is there any missing data?**

```{r}
sum(is.na(vegetables))
```
Who didn't report their data? 

```{r}
states[is.na(vegetables)]
```

Sigh. Oh Florida.

Well, at least Florida is the only state missing. Let's keep going. 


> **What is a typical value?**

```{r}
mean(vegetables)
median(vegetables)
```
Oops! Forgot to account for the missing values. Thankfully we checked earlier and saw there was only one missing value. Let's have R ignore that missing value. 

```{r}
mean(vegetables, na.rm = TRUE)
median(vegetables, na.rm = TRUE)
```

Looks like around `r round(mean(vegetables, na.rm = TRUE))`% of people in each state tend to report eating vegetables each day. Are you surprised by this, or does it seem reasonable to you? 


> **What are the smallest and largest values?**

```{r}
min(vegetables, na.rm = TRUE)
max(vegetables, na.rm = TRUE)
```

Not as much variation as we might have expected here... Interesting. 


::: {.callout-note title="Key Idea"}
Simple summary statistics answer specific questions about a variable, but no single number tells the whole story.
:::

---

## Looking at the Distribution

A histogram helps us understand the overall shape of the data.

```{r}
hist(vegetables)
```

As you look at the histogram, ask:

- What is the shape?
- Where is the center?
- How spread out are the values?
- Are there any possible outliers?

---

## Exploring a Categorical Variable

Suppose we want to understand the distribution of states across regions. As we saw earlier, this is a *categorical variable*, which R stores as text (character) data. 

```{r}
typeof(regions)
```

We could print the entire regions vector, but let's start by figuring out just how many categories there are. What are the options for region in this data set? 

```{r}
unique(regions)
```

Interesting. Many Official government data sets use more regions than this (around 9) to help account for geographic variability. With only 4 regions, I wonder how many states are in each region? 

```{r}
table(regions)
```
Seems pretty balanced among the regions, with a couple fewer states in the Northeast. What's the percent breakdown of states in each region? 

```{r}
table(regions) / length(regions)
```
So around 22% - 26% of states are classified in each of the four regions. 

---

## Looking at the Categories

This distribution is fairly equal across the categories, but it's still a good idea to visualize it with a graph: 

```{r}
barplot(table(regions))
```

Ask yourself:

- Which region has the most states?
- Which has the fewest?
- Are the counts fairly balanced?

::: {.callout-note title="Key Idea"}
Choose summaries that match the type of variable.
:::

---

## Try It

Open the companion R guide and explore the `physical` variable.

- What type of variable is this? Categorical or quantitative? 
- Is there any missing data? 
- What is the mean?
- What is the median?
- What are the minimum and maximum values?
- What does the histogram suggest about the distribution?

Remember to account for the missing values you learned about in the previous lesson.

---

## One More Question...

Suppose we now ask:

> **Does vegetable consumption differ by region?**

We know how to summarize vegetables.

We know how to summarize region.

But how do we summarize **vegetables separately for each region**, or create **one histogram for each region**?

Those questions are much harder to answer using only the base R tools we've learned so far.

---

## Looking Ahead

In the next module, we'll begin using the **tidyverse** and **ggplot2**.

These tools make it much easier to summarize data, compare groups, and create attractive graphics.

---

## Lesson Summary

- Exploratory data analysis begins by asking questions about a variable.
- Quantitative variables can be summarized with numerical summaries and histograms.
- Categorical variables can be summarized with frequency tables and bar plots.
- Different variable types require different summaries.
- More complex questions require more powerful tools.

---

## R Reference

| Function | Purpose |
|----------|---------|
| `length()` | Count observations |
| `mean()` | Calculate the mean |
| `median()` | Calculate the median |
| `min()` | Smallest value |
| `max()` | Largest value |
| `hist()` | Create a histogram |
| `unique()` | Identify categories |
| `table()` | Count categories |
| `barplot()` | Create a bar plot |
